---
title: "Class 11 Introduction to Genome Informatics Lab"
author: "Sindy"
format: pdf
editor: visual
---

## Q13:  Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
summary(expr)
table(expr$geno)
nrow(expr)
table(expr$geno)/nrow(expr)

```



## Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Boxplot

```{r}
library(ggplot2)
```


```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) + geom_boxplot(notch = TRUE)
```

> what could you infer from the relative expression value between A/A and G/G displayed in this plot?

There is a statistically significant difference between the expression levels of A/A and G/G. A/A is more abundant in the population.

> Does the SNP effect the expression of ORMDL3?

Yes. The OMIM entry for '% 611403 ASTHMA-RELATED TRAITS, SUSCEPTIBILITY TO, 6' states that being homozygous GG increases risk of asthma when exposed to environmental factors.


























# Question 6 Math

```{r}
MXL <- read.csv("MXL_asthma.csv")
head(MXL)
```


```{r}
table(MXL$Genotype..forward.strand.)


```

Proportion of G homozygous

```{r}
table(MXL$Genotype..forward.strand.)/nrow(MXL)*100
```

