---
title: "Class19_Pertussis"
author: "Sindy"
format: pdf
editor: visual
---
# 1. Investigating pertussis cases by year


## Is Pertussis on the rise?

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

Scrape the CDC data on Pertussis cases per year in the US from their website here: https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html#print 

```{r}


cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L),
  Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617)
)

```





```{r}
library(ggplot2)
ggplot(cdc, aes(Year, Cases, color = "pink")) + 
  geom_point() + geom_line() +
  labs(title = "Pertussis cases by Year in the USA")
```

# 2. A tale of two vaccines (wP & aP)

> Q2. add lines to the previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine.

```{r}
ggplot(cdc, aes(Year, Cases, color = "pink")) + 
  geom_point() + 
  geom_vline(xintercept = 1946, color = "darkgreen", linetype = 2) + geom_vline(xintercept = 1996, color = "blue", linetype = 2)
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There was an increase in cases a few years after the change in the vaccine. Perhaps the infants that were vaccinated in 1996 with the potentially less effective aP vaccine, enter school around year 2000 (when there is a small increase in cases), and with the increase contact with other children that could potentially be infected, thus leading to an increase in cases.

# 3. Exploring CMI-PB data

## The CMI-PB API returns JSON data


Load "jsonlite" package after installing it in the Console

```{r}
library(jsonlite)
```

Now let's get the data from the CMI-PB website

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)

head(subject)

```


> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```


> Q5. How many Male and Female subjects/patients are in the dataset?


```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```


# Side-Note: Working with dates

```{r}
library(lubridate)
```

Looks like two of the columns of 'subject' contain dates in the Year-Month-Day format.

```{r}
head(subject$year_of_birth, 3)
head(subject$date_of_boost, 3)
```

```{r}
today()
```

```{r}
time_length( today() - ymd("2000-01-01"), "years")
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)

```

What is the average age across all subjects?
```{r}
mean(subject$age)
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )

```
```{r}
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

The two ages seem different to me, they are about 10 years apart


> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```








> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

Histogram of ages

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) 
```
Yes, they are very different.


# Joining multiple tables


Let's read more data from CMI-PB. We will start with the Specimen table. 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE)
```

```{r}
head(specimen)
```

We want to "join" the 'subject' and 'specimen' tables to have all the metadata we need for later analysis. We can use the dplyr *_join() functions for this task.

subject is y




>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:


```{r}
meta <- inner_join(specimen, subject)
dim(meta)
```


Now read Ab titer experimental data


> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE)
head(titer)
```

I need to link a.k.a. join this ab titer data with our meta data to have all the information we need to do some actual analysis like comparing Ab levels at different times after boost, comparing aP vs wP responses over time, etc

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```
These large data sets are stored in smaller chunks to make them more accessible (downloadable) so that you can study the smaller/individual data

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

There are 6 isotypes.


> Q12. What do you notice about the number of visit 8 specimens compared to other visits?

```{r}
table(abdata$visit)
```

The number of specimens for 8 is very low, because the data is still being collected.



# 4. Examine IgG1 Ab titer levels


> Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:


We should exclude visit 8 (still ongoing data collection) and focus on one isotype. Here I pick IgG1

```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```

and a plot

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2)
```














